Concurrent Programming with CXL

### 2020-05-18 book/paper.tex
Added chapter on CXL values.  Need more examples

What is going on is that the CXL program is compiled to machine instructions,
and it is the machine instructions that are executed by the underlying
machine.  The details of this appear in the next chapter, but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

@@

@@

Instead of bits and bytes, a CXL machine manipulates \emph{CXL values}.
CXL values are recursively defined: they include booleans, integers, strings,
sets of CXL values, and \emph{dictionaries} that map CXL values to other CXL
values.

A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
a CXL.
\item Processes:  Initially there is just the main process, but any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
a program counter, a stack of CXL values,
and a single mutable general purpose \emph{register} that contains a CXL.
\end{itemize}

@@

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty tuple, \texttt{()}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
(In CXL, dictionary values are enclosed by \texttt{dict\{} and \texttt{\}}.)
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.  CXL in fact
supports this.  It has a method \texttt{atLabel($L$)}, where $L$
is the string name of a label (in this case, \texttt{"cs"}).  The
method returns a bag of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.

@@

\chapter{CXL Methods and Dictionaries}

>>>>

What is going on is that the CXL program is compiled to machine instructions,
and it is the machine instructions that are executed by the underlying
machine.  The details of this appear in the chapter~\ref{ch:cxlmachine},
but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

@@

\chapter{CXL Values}

CXL programs manipulates CXL values.
CXL values are recursively defined:
they include booleans (\texttt{False} and \texttt{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of CXL values,
and dictionaries that map CXL values to other CXL values.

Another type of CXL value is the \emph{atom}.  You have already
seen them: the names of constants, variables, and methods are
all atoms.  An atom is denoted using a period followed by the
name.  For example, the atom for the \texttt{process} method
is denoted \texttt{.process}.

CXL makes extensive use of dictionaries.  Their syntax and properties
are a little different from Python.
A dictionary maps keys to values.
Unlike Python, any CXL value can be a key, including another
dictionary.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar, but in CXL square brackets can be used in the
same way as parentheses, so you can express it in the more familiar
form:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if the key is an atom, say \texttt{.count}, then you
can write \texttt{d.count} instead of having to write
\texttt{d[.count]} (both will work).

Tuples are special forms of dictionaries where the keys are
the indexes into the tuple.  For example, the tuple
\texttt{(5, False)} is the same CXL value as
\texttt{dict\{ 0: 5, 1: False \}}.
The empty tuple \texttt{()} is the same value as \texttt{dict\{\}}.
As in Python, you can create singleton tuples by adding an
extra ",". For example, \texttt{(1,)}.

Again, square brackets and parentheses work the same in CXL, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if \texttt{x == [ False, True ]},
then \texttt{x[0] == False} and \texttt{x[1] == True}.

Dictionaries have two handy unary operators.  If \texttt{d} is a
dictionary, then \texttt{keys d} (or equivalently \texttt{keys(d)})
returns the set of keys and \texttt{len d} returns the size of
this set.

@@

Instead of bits and bytes, a CXL machine manipulates CXL values.
A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
a CXL.
\item Processes:  Initially there is just the main process, but any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
a program counter, a stack of CXL values,
and a single mutable general purpose \emph{register} that contains a CXL.
\end{itemize}

@@

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty tuple, \texttt{()}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.  CXL in fact
supports this.  It has a method \texttt{atLabel $L$}, where $L$
is the atom of the label (in this case, \texttt{.cs}).  The
method returns a bag of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.

@@

\chapter{CXL Methods and Pointers}
